\huge Why does the gut have so many species? \\ {\Large The Paradox of the Plankton in human microbial ecology}

### 2014-10-31 manuscript.tex
Finished intended text; spellchecked.

\section{Introduction}
The average, healthy human gut is home to more than one thousand species of
bacteria and archaea~\citep{Claesson2009,Brestoff2013}.
This mega-diverse microbial assemblage has been shown to contibute to human
health in numerous ways.
Fermentation in the large intestine produces a suite of \glspl*{SCFA}
which contribute to energy acquisition and regulation of
the gut epithelium~\citep{TODO}.
Both the adaptive and innate immune systems may be intially trained,
and subsequently calibrated,
by interactions with commensal bacteria in the gut~\citep{TODO}.
A healthy microbiome also appears to protect the host from certain
infections, notably \cdiff~\citep{TODO}.
A number of studies have correlated species diversity in the gut
with disease states,
including obesity~\citep{TODO},
% See HMP 2012 for this and...
\gls*{IBD}~\citep{TODO},
Crohn's disease~\citep{Manichanh2006},
and recurrent \cdiff\ infections~\citep{Chang2008}.

@@

This environmental differentiation between body sites may be
paired with microenvironmental variation at the millimeter
scale~\citep{TODO}.
Temporal fluctuations, as well, are probably a ubiquitous
experience for microbes in the human gut due to physiological
cycles of the host and directional flow of lumen contents.
Both temporal and spatial heterogeneity have been studied as
mechanisms of coexistence~\citep{Chesson2000}.
When environmental changes invert the direction of selection,
the frequency of these fluctuations (whether in space or time)
is expected to have an important effect on the outcome of
competition~\citep{Hutchinson1961}.

\begin{verbatim}
-  Are heterogeneities in the gut at the right scale to
   explain microbiome diversity?
-  Temporal heterogeneity, dormancy
-  Beta diversity
-  Intermediate disturbance~\citep{Connell1978}
    -  Original model relies on successional processes
\end{verbatim}

\subsubsection{Evolutionary Trade-offs}
The observation that the outcome of competition is context
dependent reflects another central theme of community ecology.
No species is ecologically superior in all environments,
demonstrating the universality of physiological trade-offs.
Negative pleiotropy between the ability to grow on
different substrates directly explains resource specialization.
Besides those based on environmental and trophic trade-offs,
an entire suite of coexistence mechanisms,
some of which will be described here,
depend on more abstract correlations between species traits.

\begin{verbatim}
-  Competitor/colonizer trade-off
-  Storage Effect (or at least mention non-linear effects)
\end{verbatim}

@@

One such equalizing mechanism is environmental variability.
When climatic variables fluctuate at a frequency similar to the
rate of competitive exclusion, such that the relative fitness of
two competitors is continually inverted,
fitness difference are effectively brought closer
to zero~\citep{Hutchinson1961,Chesson1981}
% TODO: Is the chesson ref correct?
Short term diversity can therefore be increased by
environmental fluctuations at the right time scale.
In the human gut, this mechanism may be highly important,
since temporal fluctuations are ubiquitous and occur at a
variety of time scales~\citep{TODO}.
Alternatively, even in static environments,
fitness equivalency may be the norm,
a hypothesis supported by the prevalence of
physiological trade-offs~\citep{TODO}.

@@

Stephen Hubbel's Unified Neutral Theory of Biodiversity~\citeyearpar{TODO}
attempted to replicate this success,
but with a more explicit treatment of mechanism
and several controversial assumptions which limit applicability.
Hubbel's model has been
reformulated for microbial populations~\citep{Sloan2005},
and may serve as a valuable null hypothesis in identifying community
compositions which cannot be explained
by purely neutral processes~\cite{VenkataramenTODO}.
A significant fraction of the species diversity in the human gut may be
transients from the external environment.
If this hypothesis is correct then community composition should reflect
primarily neutral processes.

\begin{verbatim}
-  How do they survive the stomach acid and oxygen rich atmosphere?
\end{verbatim}

>>>>

\section{Introduction}
The average, healthy human gut is home to more than one thousand species of
bacteria and archaea~\citep{Claesson2009,Brestoff2013}.
This mega-diverse microbial assemblage has been shown to contribute to human
health in numerous ways.
Fermentation in the large intestine produces a suite of \glspl*{SCFA}
which contribute to energy acquisition and regulation of
the gut epithelium~\citep{TODO}.
Both the adaptive and innate immune systems may be initially trained,
and subsequently calibrated,
by interactions with commensal bacteria in the gut~\citep{TODO}.
A healthy microbiome also appears to protect the host from certain
infections, notably \cdiff~\citep{TODO}.
A number of studies have correlated species diversity in the gut
with disease states,
including obesity~\citep{TODO},
% See HMP 2012 for this and...
\gls*{IBD}~\citep{TODO},
Crohn's disease~\citep{Manichanh2006},
and recurrent \cdiff\ infections~\citep{Chang2008}.

@@

This environmental differentiation between body sites may be
paired with microenvironmental variation at the millimeter
scale~\citep{TODO}.
Temporal fluctuations, as well, are probably a ubiquitous
experience for microbes in the human gut due to behavioral and physiological
cycles of the host and directional flow of lumen contents.
One study suggests that the gut microbiome may have
its own circadian rhythm~\citep{Thaiss2014}.
Both temporal and spatial heterogeneity have been studied as
mechanisms of coexistence~\citep{Chesson2000}.
When environmental changes invert the direction of selection,
the frequency of these fluctuations (whether in space or time)
is expected to have an important effect on the outcome of
competition~\citep{Hutchinson1961}.

In many plants and some animals, dormant or resistant life-stages can persist
in the environment for long periods of time.
These `seed banks' have important implications for local diversity,
since they integrate species richness
over past environmental conditions~\citep{TODO}.
Microbial seed banks have been shown to contribute to the diversity found in
soil~\citep{LennonTODO}.
While dormant stages are probably quickly washed out of the linear
reaches of the human gut,
the cecum and appendix may provide for the long-term persistence
of slower growing microbes~\citep{TODO}.

\subsubsection{Evolutionary Trade-offs}
The observation that the outcome of competition is context
dependent reflects another central theme of community ecology.
No species is ecologically superior in all environments,
demonstrating the universality of physiological trade-offs.
Negative pleiotropy between the ability to grow on
different substrates directly explains resource specialization.
Besides those based on environmental
and trophic trade-offs~\citep{Litchman2007},
an entire suite of coexistence mechanisms
depend on more abstract correlations between species traits.

One such mechanism is the competitor-colonizer trade-off
(\citealp{Levins1971}; for a brief review see \citealp{Yu2001}),
which results in coexistence through the partitioning of space between species.
Given a trade-off between dispersal or colonization ability
and competitive ability,
a superior colonizer can quickly occupy habitat space which has become open
due to local catastrophe or chance extinction.
While the superior competitor will eventually reach the same locality
and outcompete the first species,
with a high enough density of unoccupied space
the colonizer may be able to disperse quickly enough to maintain a stable
population.
Depending on the particular parameters,
these `fugitive' species can coexist indefinitely with the superior competitor.
A similar model may describe successional processes
on food particles in the gut~\citep{TODO}.

@@

One such equalizing mechanism is environmental variability.
When climatic variables fluctuate at a frequency similar to the
rate of competitive exclusion, such that the relative fitness of
two competitors is continually inverted,
fitness difference are effectively brought closer
to zero~\citep{Hutchinson1961,Chesson1981}
% TODO: Is the Chesson ref correct?
Short term diversity can therefore be increased by
environmental fluctuations at the right time scale.
In the human gut, this mechanism may be highly important,
since temporal fluctuations are ubiquitous and occur at a
variety of time scales~\citep{TODO}.
Alternatively, even in static environments,
fitness equivalency may be the norm,
a hypothesis supported by the prevalence of
physiological trade-offs~\citep{TODO}.

@@

Stephen Hubbel's Unified Neutral Theory of Biodiversity~\citeyearpar{TODO}
attempted to replicate this success,
but with a more explicit treatment of mechanism
and several controversial assumptions which limit applicability.
Hubbel's model has been
reformulated for microbial populations~\citep{Sloan2005},
and may serve as a valuable null hypothesis in identifying community
compositions which cannot be explained
by purely neutral processes~\cite{VenkataramenTODO}.
It is possible that
a significant fraction of the species diversity in the human gut may be
transients from the external environment.
If this hypothesis is correct then community composition should reflect
primarily neutral processes.
